Automated inference of production rules for glycans

### 2021-07-04 papers/sugar-synth/algo.tex
dsadsa

In this section, we present a method to solve the synthesis problem of finding
a set of production rules that produce a given set $\mu$ of molecules.
Our synthesis method \textsc{SugarSynth} is presented in Algorithm~\ref{alg:sugar-synth}.
Here, we are considering only the single compartment case.
The generalizations are discussed in Appendix~\ref{sec:variations}.

@@

A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations.
The first situation is when a node position is in the expanding part.
The dark gray nodes are the expanding nodes.
The second situation is when a node position is not in the rule.
The dashed area are the absent nodes. Among the matching nodes, we have two cases.
The third situation is when a node position is in the matching part
and has expanding descendants.
The nodes on the solid path from $v_0$ to the root $v_e$ of the expanding part are in the third situation.
% the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is the rest of the node positions
in the matching part, which is the light gray area.
% The nodes in the rest of the light gray area are in the fourth situation.
A variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the unbounded set of variables with domain $K$. 
% We will use the pool of $KVars$ to add variables to the templates.
The templates are defined as follows.

@@

\input{alg-encode-produce.tex} 
\vspace{-3mm}
Algorithm~\ref{alg:produce} presents function $\textsc{EncodeProduce}$
that returns the encoding.
It takes a molecule $m$ and a set of rules $T$.
Both the inputs can be template or concrete.
% The function is tolerant of the variations, but
Our presentation assumes that the molecule is concrete and
the rules are templates.
This will cover the case when the $\textsc{EncodeProduce}$ is called at
line~\ref{line:molenc} in Algorithm~\ref{alg:sugar-synth}.
However, at line \ref{line:consNewR} the molecule is a template
and the rules are concrete, which we will discuss later.
% We will discuss the handling of this situation at the end of this subsection.
$\textsc{EncodeProduce}$ uses the help of three other supporting functions,
$\textsc{EncodeP}$, $\textsc{MatchTree}$, and $\textsc{MatchCut}$.
% All of them are presented in algorithm~\ref{alg:produce}.

@@

@@

Our algorithm always terminates. At each iteration, at least one solution, i.e,
a set of rules is discarded.
In fact, we discard many because in each iteration we reject production of a counterexample molecule, but this may discard whole range of rule sets. 
Since the set of all possible rules is finite, we
guarantee termination. In the worst case, the entire search space is explored. For a problem having 10 node types, 3 as maximum rule depth, 10 rules to be synthesized, 3 compartments
and fast-slow reactions, the search space can only be as large as $\approx 10^{74}$ rules $(2^{10}*{(10+3-1) \choose (3-1)}*10^{(2^3 - 1)*10})$.
Our templates are sufficiently expressive to cover all aspects of biology.
The viability of the synthesized rules can only be verified by conducting wet experiments.

>>>>

In this section, we present a method to solve the synthesis problem of finding
a set of production rules that produce a given set $\mu$ of molecules.
Our synthesis method \textsc{SugarSynth} is presented in Algorithm~\ref{alg:sugar-synth}.
Here, we are considering only the single compartment case.
The generalizations are discussed in
\choosefinal{extended version~\cite{arxiv-this}.}{ Appendix~\ref{sec:variations}.}

@@

A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations.
The first situation is when a node position is in the expanding part.
The dark gray nodes are the expanding nodes.
The second situation is when a node position is not in the rule.
The dashed area are the absent nodes. Among the matching nodes, we have two cases.
The third situation is when a node position is in the matching part
and has expanding descendants.
The nodes on the solid path from $v_0$ to the root $v_e$ of the expanding part are in the third situation.
% the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is the rest of the node positions
in the matching part, which is the light gray area.
% The nodes in the rest of the light gray area are in the fourth situation.
A variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the unbounded set of variables with domain $K$. 
% We will use the pool of $KVars$ to add variables to the templates.
Our templates are sufficiently expressive to cover all aspects of biology.
The templates are defined as follows.

@@

Algorithm~\ref{alg:produce} presents function $\textsc{EncodeProduce}$
that returns the encoding.
It takes a molecule $m$ and a set of rules $T$.
Both the inputs can be template or concrete.
% The function is tolerant of the variations, but
Our presentation assumes that the molecule is concrete and
the rules are templates.
This will cover the case when the $\textsc{EncodeProduce}$ is called at
line~\ref{line:molenc} in Algorithm~\ref{alg:sugar-synth}.
However, at line \ref{line:consNewR} the molecule is a template
and the rules are concrete, which we will discuss later.
% We will discuss the handling of this situation at the end of this subsection.
$\textsc{EncodeProduce}$ uses the help of three other supporting functions,
$\textsc{EncodeP}$, $\textsc{MatchTree}$, and $\textsc{MatchCut}$.
% All of them are presented in algorithm~\ref{alg:produce}.

@@

\vspace{-3mm}
\input{alg-encode-produce.tex}

@@

\begin{theorem}
  If \textsc{SugarSynth}($\mu$, $d$, $n$) returns a set of rules $R$, then if $R$ produces a molecule that has depth less than $h$, the it is in $\mu$ (soundness).
  If it fails to synthesize rules, there are no rule set from the budget of depth $d$ and number $n$ (completeness).
\end{theorem}

The soundness is true due to the construction of constraints.
Our algorithm always terminates. At each iteration, at least one solution, i.e,
a set of rules is discarded.
In fact, we discard many because in each iteration we reject production of a counterexample molecule, but this may discard whole range of rule sets. 
Since the set of all possible rules is finite, we
guarantee termination.
Therefore, the completeness holds.
In the worst case, the entire search space is explored.

### 2021-07-04 papers/sugar-synth/appendix.tex
dsadsa

\documentclass{llncs}
%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

% \documentclass[conference,10pt]{IEEEtran}

% \bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
% \citestyle{acmauthoryear}   %% For author/year citations

\input{macro.tex}
\setlength{\textfloatsep}{0pt}%

\title{Automated inference of production rules for glycans}
\author{Ansuman Biswas$^2$, Ashutosh Gupta$^1$, Meghana Missula$^1$, and Mukund Thattai$^2$}
\institute{$^1$IITB and $^2$NCBS}

\date{May 2019}

\begin{document}

\maketitle

\begin{abstract}
  Glycans are tree-like polymers made up of sugar monomer building blocks. They are found on the surface of all living cells, and distinct glycan trees act as identity markers for distinct cell types. Proteins called GTase enzymes assemble glycans via the successive addition of monomer building blocks. The rules by which the enzymes operate are not fully understood. In this paper, we present the first SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans from a cell. We have built a tool based on the method and applied it to infer rules based on published glycan data.
\end{abstract}

\section{Introduction}
\label{sec:intro}
\input{intro}

\section{Production of Glycans}
\label{sec:bio}
\input{bio.tex}

% \clearpage

\section{Motivating example}
\label{sec:motivation}
\input{motivation.tex}

\section{Modelling of the synthesis problem}
\label{sec:model}
\input{model.tex}

% \clearpage

\section{Method for the synthesis problem}
\label{sec:algo}
\input{algo.tex}

\section{Experiments}
\label{sec:experiments}
\input{experiments.tex}

% \section{Related work}
% \label{sec:related}
% \input{related.tex}

\section{Conclusion and future work}
% ##limitations in the model because of limited information,
% a)the boundary between two compartments is fuzzy, 
% b) the growths of two branches are correlated/ not correlated..and some more points like these, for which wealth of data is not available, hence the model was unable to incorporate the specifics
\label{sec:conclusion}
\input{conclusion.tex}
\nocite{Charles2013}
\bibliographystyle{unsrt}
\bibliography{biblio}

\appendix

@@

% 
% To resolve this paradox we borrow from the field of
% al-gorithmic self-assembly, which explores how small buildingblocks
% with stochastic local interactions assemble into various global
% configurations [23]–[25].  A central inverse problemin self-assembly
% is to design building blocks that assembleinto a target shape.  This
% process has been studied using thetheoretical framework of Wang tiles:
% square building blockswith colored sides that stick to one another
% along sides withmatching colors [26].  For efficient assembly of Wang
% tiles,their coloring must be cleverly chosen so the target shape isthe
% unique terminal output of the stochastic assembly process[27].   That
% is, their assembly must be algorithmic.  Glycans may be considered a
% natural realization of the Wang construct,with monomers acting like
% tiles whose stickiness is encodedby GTase enzymes.  If glycan
% biosynthesis could be madealgorithmic, cells could suppress unwanted
% byproducts andgenerate only desired glycan oligomers with high yield.

\end{document}

>>>>

@@

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2021-07-04 papers/sugar-synth/intro.tex
dsadsa

A given cell produces a specific set of glycan molecules that are present in the cell. Since different cells produce different sets of molecules,
the assembly process must be programmable: the assembly process includes a set of production rules. The reactions that underlie glycan production are carried out by enzymes known as GTases \cite{Varki2017}. There are hundreds of enzymes present in a given cell: each enzyme is a protein encoded by a distinct gene, which carries out a distinct biochemical reaction. The enzymes thus execute the production rules. A glycan tree is assembled piece by piece in successive steps. At each step, a production rule adds a small piece of a tree %(a monomer)
at the leaves or internal nodes of the current tree.
Not all the rules are applicable at all the leaves. The monomer at a leaf and current surroundings of the leaf controls
the applicability of a rule on the leaf.
Identifying the exact set of the production rules by extensive
biochemical experiments are costly and often needs an initial
hypothesis for the rules to test. Biologists must identify the production rules and their control
conditions by manually analyzing the set of observed glycans in a cell
and using prior knowledge of biochemistry.
We estimate that there are approximately $10^{70}$
possible rule sets if we consider all biological variations.
This is an error-prone process since the production rules must generate
exactly the set of molecules in the cells and nothing else; and moreover, the data sets about which glycans are present in which cells are often incomplete. Manually comprehending all possible tree generation rules is difficult and ad-hoc. It would be useful to automate the process of learning which rules are operating in a given cell, based on incomplete data.

@@

Our encoding to constraints depends on the model of execution of the rules.
The current biological information is not sufficient to make a precise
and definite model, and do the synthesis.
We have also explored the variations of the models. For example, all rules
may apply simultaneously. They apply in batches because they
stay in different compartments.
The molecule under assembly may flow through the compartments.
The distribution of the stay of the molecules in a compartment also affects
the execution model.
Furthermore, we may have variations in the type and quality of data available to us.
We have added support for the variations in our method.

% \todo{Related work}
We have implemented the method in our tool~\ourtool. The tool takes input data in its own format.
We have applied the tool on data sets from published sources (these are available in the database UniCarbKB \cite{Campbell2013}).
% We have obtained the rules for the data sets.
The output rules are within the expectations of biological
intuition.

The key contributions of this paper are the following.
\begin{itemize}
\item We identified {\em a new area} of application for formal methods in biology.
\item We have developed a {\em novel} synthesis method for discovering the production rules of glycan molecules from the output of the rules.
\item We have implemented the method in a tool and applied it to data sets.
\end{itemize}

We organize the paper as follows.
In Section~\ref{sec:bio}, we introduce the biological background.
%of the synthesis problem and present the basis of the formal model.
% If the reader wants to avoid getting into the biological details, she may skip the section.
In Section~\ref{sec:motivation}, we present a motivating example to illustrate our method.
In Section~\ref{sec:model}, we present the formal model of the glycans
and their production rules.
In Section~\ref{sec:algo}, we present our method for the synthesis problem.
%and the modifications in our methods to support the variations.
In Section~\ref{sec:experiments}, we present our implementation  and experiments. %on some glycan data.
In Section~\ref{sec:conclusion}, we conclude the paper.
We present the extended related work in Appendix~\ref{sec:related}.

>>>>

A given cell produces a specific set of glycan molecules that are present in the cell. Since different cells produce different sets of molecules,
the assembly process must be programmable: the assembly process includes a set of production rules. The reactions that underlie glycan production are carried out by enzymes known as GTases \cite{Varki2017}. There are hundreds of enzymes present in a given cell: each enzyme is a protein encoded by a distinct gene, which carries out a distinct biochemical reaction. The enzymes thus execute the production rules. A glycan tree is assembled piece by piece in successive steps. At each step, a production rule adds a small piece of a tree %(a monomer)
at the leaves or internal nodes of the current tree.
Not all the rules are applicable at all the leaves. The monomer at a leaf and current surroundings of the leaf controls
the applicability of a rule on the leaf.
Identifying the exact set of the production rules by extensive
biochemical experiments are costly and often needs an initial
hypothesis for the rules to test. Biologists must identify the production rules and their control
conditions by manually analyzing the set of observed glycans in a cell
and using prior knowledge of biochemistry.
We estimate that there may be more then $10^{70}$
possible rule sets if we consider all biological variations.
\footnote{For a problem having 10 monomers, 10 rules, 3 as rule size, 3 compartments
and fast-slow reactions, the search space is $\approx 10^{74}$ rules $(2^{10}*{(10+3-1) \choose (3-1)}*10^{(2^3 - 1)*10})$.}
This is an error-prone process since the production rules must generate
exactly the set of molecules in the cells and nothing else; and moreover, the data sets about which glycans are present in which cells are often incomplete. Manually comprehending all possible tree generation rules is difficult and ad-hoc. It would be useful to automate the process of learning which rules are operating in a given cell, based on incomplete data.

@@

Our encoding to constraints depends on the model of execution of the rules.
The current biological information is not sufficient to make a precise
and definite model, and do the synthesis.
We have also explored the variations of the models. For example, all rules
may apply simultaneously. They apply in batches because they
stay in different compartments.
The molecule under assembly may flow through the compartments.
The distribution of the stay of the molecules in a compartment also affects
the execution model.
Furthermore, we may have variations in the type and quality of data available to us.
\choosefinal{We support the variations, which is presented in 
  the extended version of this paper is available at~\cite{arxiv-this}.}{We support the variations.}

% \todo{Related work}
We have implemented the method in our tool~\ourtool. %The tool takes input data in its own format.
We have applied the tool on data sets from published sources (available in the database UniCarbKB \cite{Campbell2013}).
% We have obtained the rules for the data sets.
The output rules are within the expectations of biological
intuition.

% The key contributions of this paper are the following.
% \begin{itemize}
% \item We identified {\em a new area} of application for formal methods in biology.
% \item We have developed a {\em novel} synthesis method for discovering the production rules of glycan molecules from the output of the rules.
% \item We have implemented the method in a tool and applied it to data sets.
% \end{itemize}

We organize the paper as follows.
In Section~\ref{sec:bio}, we introduce the biological background.
%of the synthesis problem and present the basis of the formal model.
% If the reader wants to avoid getting into the biological details, she may skip the section.
In Section~\ref{sec:motivation}, we present a motivating example to illustrate our method.
In Section~\ref{sec:model}, we present the formal model of the glycans
and their production rules.
In Section~\ref{sec:algo}, we present our method for the synthesis problem.
%and the modifications in our methods to support the variations.
In Section~\ref{sec:experiments}, we present our implementation  and experiments. %on some glycan data.
In Section~\ref{sec:conclusion}, we conclude the paper.
\choosefinal{}{We present the extended related work in Appendix~\ref{sec:related}.}

### 2021-07-04 papers/sugar-synth/main.tex
dsadsa

\section{Extended related work}
\label{sec:related}

% Success of formal methods
Formal methods have successful in applications to a vast range of
problems from the analysis of systems.
%
In particular,
the verification of hardware designs
using SAT solver based model-checking has been applied
to industrial-scale problems~\cite{biere1999symbolic2}.
%
The verification of software systems is a much harder problem, and
many methods like bounded model-checking~\cite{biere2003bounded},
abstract interpretation~\cite{lattice77}, and
counterexample guided abstraction refinement (CEGAR)~\cite{ClarkeCEGAR} have
been successful for the problem.
%

Synthesis of programs that implements a given behavior
has been a focus of research for a while~\cite{PnueliSynthesis}.
%
In recent years, the synthesis of programs using SAT/SMT solvers
has gained momentum.
%
The approach encodes the search of a program that exhibits a certain
behavior into a satisfiability problem.
%
The solvers attempt to find a solution to the satisfiability problem.
%
The solution is the synthesized program.
%
In~\cite{SrivastavaSynthesis,Solar-Lezama2005},
the method has been successfully applied
to find programs that satisfy quantified specifications like sorting
programs, and have missing ``holes'' and need proper implementations
for them.
%
In~\cite{exampleSynth},
a set of pairs of input and output examples is the specification of a program,
and the synthesis method searches for programs in a space defined by a
template~\cite{sygus}.

% Boolean networks

% Program synthesis
Since the formal methods have been exhibiting effectiveness in the vast
range of problems, the
systems biology community has been applying the methods for various
biological problems~\cite{fisher2007executable}.
The formal methods approach is distinctly different from the statistical
approach traditionally used in biology.
In formal methods, we aim to match precisely the expected behavior rather
than learn approximate artifacts from biological data.
The use of formal methods belongs to two broad categories:
the analysis of biological models and the synthesis of the models.

The key focus has been to model gene regulatory networks (GRNs), since
they are the core of the central dogma of biology.
Boolean networks are often used to model
GRNs to find stable points or attractors, i.e., nodes in the networks
where a system eventually reaches no matter where it starts~\cite{wang2012BooleanOverview}.
% In~\cite{boolean-networks-stabilty}, the networks are used to evaluate
% stability of GRNs under mutations.
The Boolean networks are not often sufficient to adequately model the behavior
of GRNs.
The continuous-time Markov chains (CTMCs) are used to bring in the aspects of
timing and probabilistic constraints to GRNs.
The transient behaviors of GRNs are also crucial for various aspects
of biology.
% For example, the decision making in some cells is based on
% response timing to stimulus~\cite{response-timing-paper}.
In~\cite{delayedCTMC}, a method based on
model-checking is used to estimate the time evolution of the probability
distributions over states of GRNs.

There are many biological processes where we can observe the system behavior,
but the exact mechanisms of the processes are not known.
%
Recently, the methods for formal synthesis are finding their applications in biology
\cite{dunn2014defining,xuPluripotency,booleanModelKarp13,paoletti2014analyzing}.
%
In~\cite{koksal2013synthesis},
GRNs with unknown interactions are modeled using
an automata-like model with missing components
and the specification is the variations of the behavior of the system under mutations.
Their approach uses a counterexample-guided inductive synthesis (CEGIS) based algorithm~\cite{cegis}.
CEGIS is a framework for synthesis. It first finds a program that may satisfy `samples' from
the specification. If the synthesized program satisfies the specifications, the CEGIS terminates.
Otherwise, the method learns a new sample where the program violates the specifications.
It adds the sample to the set and goes to the next iteration.
Typically, the constraint solvers find the programs during intermediate steps of CEGIS by
solving the generated queries.
%
Our method follows a similar pattern, where
a set of sampled observations is a specification, i.e., output molecules,
and we need to find the governing programs, i.e., production rules.
%
In a subsequent work~\cite{fisher2015synthesising}, a Boolean network model is used,
the functions attached to the nodes are considered unknown, and
a different kind of data sets provides the behavior specification.
In this approach, their earlier method is adopted to work in the new situation.
%

As far as we know, there has been no similar computation based analysis of glycan production rules.
%
However, there has been a theoretical analysis of the properties of the production
rules in~\cite{Jaiman2018}. The work identifies the conditions for the production
of a finite set of molecules and the cases of unique production methods.
%
In our work, we are taking the computational approach of synthesis from formal methods
instead of looking for the theoretical conditions.

\section{Variations of the synthesis problem}
\label{sec:variations}
\input{variations.tex}

\section{Experiments with variations of the synthesis problem}
\label{sec:ex-variants}
We conducted separate experiments for the variations of the synthesis problem - non-monotonic rules and unbounded molecules using synthetic data.
%
\begin{table}
  \centering
  \begin{tabular}{|c|c|c|c|c|c|c|}\hline
     & \#molecules& \#Rules & Rule depth & \#Compartments & success? & Time (in secs.) 
          \\\hline
    %%cf-mucin.sugar%%
         &   & 6  & 2 & 1 & Yes &  2.85 \\\cline{3-7}
    N1   & 6 & 6  & 2 & 5 & Yes & {\bf 1.07}  \\\cline{3-7}
         &   & 5  & 3 & 1 & No & 0.81  \\\hline

    %%horse-cg.sugar%%
         &   & 6  & 2 & 1 & Yes & 1.02  \\\cline{3-7}
    N2   & 6 & 6  & 2 & 5 & Yes & {\bf 0.76}  \\\cline{3-7}
         &   & 6  & 4 & 1 & Yes & 0.82 \\\hline

    %%sars-cov2.sugar%%     
         &   & 9  & 2  & 1  & Yes & 3.18  \\\cline{3-7}
    N3   & 9 & 8  & 2 & 1 & Yes & {\bf 2.39}  \\\cline{3-7}
         &   & 8  & 3 & 2 & Yes  &  16.69 \\\hline

    %%human-cg.sugar%%     
         &   & 6  & 3  & 1  & Yes & 73.61  \\\cline{3-7}
    R1   & 5 & 5  & 3 & 2 & Yes & {\bf 47.72}  \\\cline{3-7}
         &   & 4  & 4 & 1 & No  &  16.99 \\\hline

    % %%D4.sugar%     
         &   & 6  & 3 & 3 & Yes &  5.25 \\\cline{3-7}
    R2   & 8 & 4  & 3 & 2 & Yes & {\bf 3.44}  \\\cline{3-7}
         &   & 4  & 3 & 5 & Yes  &  4.70 \\\hline

    % %%D5.sugar%     
         &   & 7  & 3 & 2 & Yes &  6.85 \\\cline{3-7}
    R3   & 8 & 7  & 3 & 4 & Yes & {\bf 2.09}  \\\cline{3-7}
         &   & 6  & 3 & 1 & No  & 0.77  \\\hline
  \end{tabular}   
  \end{table}
 We applied our tool on the synthetic datasets N1, N2 and N3 which required some rules to have $\textsc{HardEnds}$. There is a clear trend of increasing time as the size of our dataset increases. Between N1 and N2 which have the same number of molecules in the dataset, N1 has bulkier molecules and hence the time taken is higher. Reducing the number of compartments increases the time slightly however is successful in both N1 and N2. By Reducing the number of rules to be learnt, the tool fails in N1 as it required a minimum of 6 rules. Rules produced in these datasets have $\textsc{HardEnds}$ in the appropriate places to ensure that the exact set of molecules can be made.
 %

  To demonstrate the working of our tool in case of runaway reactions or datasets with unbounded molecules, we applied our tool on the synthetic datasets R1, R2 and R3. The molecules in these datasets have repeating patterns and we observe the synthesized rules to capture this pattern. The constraints added as part of unbounded molecules take care not to declare the molecules created by the repeated application of these rules to be negative. The time taken for synthesizing the rules is particularly high for R1 as the molecules in the dataset are relatively bulky. Increasing the number of compartments or Increasing the number of rules to learn increases the time in R2. The search in the combinatorial space is expensive, hence the time taken is more.

\section{Implementation optimizations}

% \subsection{Optimizations}
We have incorporated several optimizations to gain efficiency or reduce
the number of iterations in the method. Let us discuss some of the remarkable
optimizations.

\paragraph{Ordered templates}
Our method constructs a list of templates.
Since the templates are symmetric, there is a potential for unnecessary
iterations.
Let us suppose the method rejects a set of learned rules.
In the subsequent queries include the constraints that reject the set of rules,
the solvers may choose the same set of rules again by permuting the mapping
from the templates to the learned rule.
To avoid these iterations, we define an arbitrary total order over the rules.
We add ordering constraints stating that any assignments to the list of templates should produce
ordered rules with respect to the total order.

\paragraph{Quantifier instantiations}
The constraints for rejecting the counterexample molecule is universally quantified at line~\ref{line:negCons} in Algorithm~\ref{alg:sugar-synth}.
For any assignment, the solver needs to try sufficiently many instantiations of the quantifiers
to ensure that the query is not unsatisfiable.
The instantiations sufficiently slow down the solving process.
We assist the solver by also adding an instantiation of the quantifiers that is equal
to the values of the corresponding variables in the assignment $a$ at line~\ref{line:negModel}.
We observe that for some inputs only adding the instantiations and not the universally quantified
formula is more effective.

\paragraph{Constraints for counterexample molecules}
In the presentation of \textsc{SugarSynth}, we construct constraints of counterexample molecules
at line~\ref{line:consNewR} in each iteration.
The repeated work is impractical because the formula management system of Z3 will be overwhelmed by
the construction of many terms repeatedly. In our implementation, we construct the constraints
once outside the while loop.
We pass the templates as the second parameter instead of concrete rules to~\textsc{EncodeProduce}.
Later at the solving time in line~\ref{line:negModel}, constraints are added to assign values
of the template variables that were returned by the solver at line~\ref{line:posModel}.
Due to the assignments, the templates become concrete rules in the context of solving.

\section{Features to support in future}

\subsection{Full organism data vs single cell data}

The experiments that observe the set of glycan molecules are of two kinds. In one kind, we isolate a single cell type organism and identify all the glycans present in the cells. In our presentation, we have assumed this kind of source of data. However, the experiments are difficult to conduct.
In another and more convenient way, we smash the whole organism and identify all the glycans present in all the cells.
So the information that which glycans are coming from which
cells is unknown.
In this situation, the synthesis has another task to map the molecules
in $\mu$ to several cell types.
A glycan may be present in multiple cell types.

Our method is easily adapted to consider this kind of data.
We search for a small covering set of subsets of
$\powerset{\mu}$ such that each set in the cover allows a small
set of production rules.
The cover sets indicate the different types of cells in terms of
the presence of glycans.
This variation makes the problem particularly hard.
There can be a potentially large number of covering subsets.
So far, we have not encountered a large enough data set such that
we apply the variation effectively.

\subsection{Incomplete data}

The experiments are imperfect. They may not detect all possible glycans
in a cell.
We may need to leave the possibility of allowing a few more molecules to be
producible beyond $\mu$.
We replace the no extra molecule constraint by a constraint that allows molecules
that are `similar' to molecules in $\mu$.
However, we could not imagine any measure of similarity that naturally stems from
biological intuition.
For now, our tool supports the count of the number of monomer differences
as a measure of similarity.

@@

>>>>

@@

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
